\section{Koopman Operator Theory}















\subsubsection{Hankel-DMD}
A variant of DMD is Hankel-DMD. Instead of just considering the transformation of the current state, the Hankel-DMD
approach includes time-delayed measurements \parencite{bruntonDataDrivenDynamicalSystems2019}. This allows the Koopman transformation to include longer term memory
effects. 

\TODO{Sparse identification of koopman eigenfunctions

motivation for identifying eigenfunctions rather than entire matrix:

- Identifying regression models based on nonlinear measurements will
generally result in closure issues \parencite{luschDeepLearningUniversal2018}\\
- Resulting models are often high dimension, can become uninterpretable

- In practice obtaining eigenfunctions is difficult
}



\subsection{Koopman theory for stochastic systems}
The above sections discuss the Koopman operator for deterministic systems. If a stochastic dynamical system is to be
considered however, the formulation can be adjusted. This is given by \textcite{crnjaric-zicKoopmanOperatorSpectrum2020}
in the following form: 

\begin{equation}
    \mathcal{K}g(\mathbf{x}) = \mathbb{E}[g(\mathbf{F}(\mathbf{x}))]
\end{equation}

\todo{check equation}

The eigenvalue formulation can then analogously be formulated:

\begin{equation}
    \mathcal{K}\varphi(\mathbf{x}) = \lambda^S(t)\varphi(\mathbf{x})
\end{equation}

where $\varphi$ are the eigenfunctions of the stochastic Koopman Operator and $\lambda^S(t)$ are the stochastic Koopman
eigenvalues. The stochastic Koopman operator tracks how the expected state of the stochastic dynamical system develops
in measurement space. It does not however track the variance of the the state.

\textcite{crnjaric-zicKoopmanOperatorSpectrum2020} discusses two approactes for determining the stochastic Koopman
operator. Both are based on the DMD approach, applied to the expected value of the measured state rather than on
deterministic tracjectories. The first is a modification of the regular DMD approach, while the second modifies the
Hankel-DMD approach. These methods do not inherintly consider the probability distrubution of the resulting
trajectories however.


\todo{Update text on VAMP}
Both VAMP \parencite{mardtVAMPnetsDeepLearning2018} and \textcite{zhaoDatadrivenProbabilityDensity2023} aim to approximate the Koopman operator using time-lagged covariance matrices derived from observed data. VAMP employs a variational principle to optimize the approximation within a chosen feature space, formulating the problem through the maximization of a score function based on covariance and cross-covariance matrices. \textcite{zhaoDatadrivenProbabilityDensity2023} adopt a conceptually similar framework but introduce alternative notation and implementation details, while retaining the core idea of leveraging covariance structures to estimate the operator. In essence, both methods share the same mathematical foundation, differing primarily in presentation and specific optimization strategies.

% https://scholar.google.com/scholar?q=related:cdgg75wUJSYJ:scholar.google.com/&scioq=vampnet&hl=en&as_sdt=0,5&inst=6173373803492361994



