\chapter{Uncertainty Quantification - 8p}
\label{chapter:uncertainty_quantification}



This chapter will investigate the fundamentals of uncertainty quantification. Techniques and methods relevant to the
prediction of DP performance will be identified and documented.

\section{What is Uncertainty Quantification? - 1p}
A typical uncertainty quantification problem involves a mathematical model, subject to
uncertainty about the correct parameter values \parencite{sullivanIntroductionUncertaintyQuantification2015}.
\textcite{higdonUncertaintyQuantificationError2009} define uncertainty quantification as follows:

\begin{quotation}
    "Uncertainty quantification (UQ) studies all sources of error and uncertainty, including the following: systematic
    and stochastic measurement error; ignorance; limitations of theoretical models; limitations of numerical
    representations of those models; limitations of the accuracy and reliability of computations, approximations, and
    algorithms; and human error. A more precise definition is UQ is the end-to-end study of the reliability of
    scientific inferences"
    \parencite{higdonUncertaintyQuantificationError2009}.
\end{quotation}

In general in the study of UQ, one is interested in the relationship between the input and output. Two general examples
are the forward and the inverse problem \parencite{sullivanIntroductionUncertaintyQuantification2015}. In both examples
we will consider a general model of a physical system:

\begin{equation}
y = f(x)
\end{equation}

In the forward problem, \(f, \, x \) are known. \(y\) can be calculated. However, if \(f\) is an imperfect model or 
\(x\) is subject to some uncertainty, then this uncertainty will in some way affect \(y\). This is known as the 
\textit{forward propegation of uncertainty}.

In the inverse problem \(y, \, x\) are measured. We are interested in determining \(f\). These measurements are subject to noise and as \(f\) is a model of a
real system and thus has infinitely many degrees of freedom, the problem is undetermined. 

For the considered case of a DP system, the forward problem is primarily of relevance, as the dynamics of the system are
known and the effect of uncertainty in the model and inputs is of relevance to the forecasted behavior.

\section{Types of uncertainty - 1p}
In general, sources can of uncertainty can be divided into two categories, based on their source. These are aleatoric
uncertainty and epistemic uncertainty \parencite{sullivanIntroductionUncertaintyQuantification2015}. Within any given system, these two forms of uncertainty are not mutually exclusive
\parencite{foxDistinguishingTwoDimensions2011}. 

Aleatoric uncertainty refers to uncertainty inherent in a system. It is irreducable, regardless of the accuracy of the
model or amount of training data. Pure aleatoric uncertainty is interpreted as the relative frequency of observations in
a class of possible outcomes \parencite{foxDistinguishingTwoDimensions2011}. It is attributed to stochastic behaviour. An example of aleatoric uncertainty could be noise in a measurement sensor. 

Epistemic uncertainty is uncertainty due of a lack of knowledge. Pure epistemic uncertainty can be seen as the
confidence one has in a single observation.
This can be due to uncertainty in parameter values or
in the accuracy of the model. Epistemic uncertainty can be improved by obtaining more data or using a higher fidelity
model. 

When modelling a dynamic positioning system, both forms of uncertainty are present. The system is subject to stochastic
wave and wind loading, as well as noise in sensor measurements. These are sources of aleatoric uncertainty. The fact
that the system is being modelled means epistemic uncertainty is being introduced, as no model will fully capture every
part of the system with complete accuracy. Additional epistemic uncertainty is added in the model parameters, as factors
such as hydrodynamic coefficients are unlikely to be completely accurate at any given time.  

\section{Spectral methods - 2p}
Spectral methods are numerical techniques that represent stochastic processes as series expansions using an orthogonal
basis of functions. They decompose random variables into components, similar to Fourier analysis
\parencite{heuvelineHybridGeneralizedPolynomial2014}. Polynomial Chaos Expansions (PCE) are a common spectral method for
uncertainty quantification.

\subsection{Generalised Polynomial chaos}
PCE uses polynomials as the orthogonal basis. Originally developed to model Wiener processes using Hermite polynomials, it was extended by
\textcite{xiuWienerAskeyPolynomial2002} to use other polynomials and thus represent other probability distributions. The
generalised polynomial chaos expansion (gPC) works as follows:

Let \(X\) be a random variable in probability space \(\left(\Omega, \mathcal{F}, \mathbb{P}\right)\), where \(\Omega\) is
the sample space with \(\omega \in \Omega\). \(X\) can then be formulated as follows
\parencite{heuvelineHybridGeneralizedPolynomial2014}:

\begin{equation}
\label{eq:gPC_explain_X}
X(\omega) = \sum_{i = 0}^{\infty}c_i\boldsymbol{\Psi}_i(\boldsymbol{\xi})
\end{equation}

\(\boldsymbol{\xi}\) is a random variable subject to some probability distribution.\({\boldsymbol{\Psi}}\) is a set of orthogonal polynomials that describes the
distribution of the random variable. \(c_i\) is a weight given to each term of the polynomial. These are determined by
pointwise sampling at a given point in time. Given a time dependent process \(X = X(t, \omega)\), \(c_i\) has to be
re-evaluated at each timestep. 

infinite sums are not possible in reality, so the sum is truncated at point \(M\). This gives an approximation of \(X\)
defined as:

\begin{equation}
X^M(t, \omega) \coloneq \sum_{i = 0}^{M}c_i(t)\boldsymbol{\Psi_i}(\boldsymbol{\xi})
\end{equation}
Where M is given by:
\begin{equation}
M = \frac{(P + L)!}{P!L!} - 1
\end{equation}

The truncation order \(M\) is dependent on two discretisation parameters, \(L\) and \(P\). \(P\) is the maximal polynomial order
used, and \(L\) is the dimension of the vector of random variables \(\boldsymbol{\xi} = (\xi_1, \dots, \xi_L)\). The value of M
grows rapidly as the dimensionality of the problem increases. The mean and variance of the result can be determined from
the resulting expansion. 

\begin{align}
    \overbar{X}(t) &= c_0(t) \\
    \mathit{Var} \, \left[ X(t) \right] &= \sum_{i = 1}^{M} c_i(t)^2
\end{align}

Overall this method works more efficiently than Monte-Carlo as the coefficients for the sum can be estimated using far
fewer model evaluations. Additionally, once the polynomial surrogate model is constructed, it can be evaluated quickly
for any variations of the input values. 

\subsection{Time dependent generalised polynomial chaos}
\textcite{gerritsmaTimedependentGeneralizedPolynomial2010} introduced the concept of time-dependent generalised
polynomial chaos (TD-gPC). One of the limitaions of gPC is a lack of convergence when the dependence on \(\boldsymbol{\xi}\) is
strongly non-linear. This is especially the case when considering long term integration
\parencite{heuvelineHybridGeneralizedPolynomial2014}. The difference between this method and the time independent method
is that with TD-gPC, the orthogonal polynomials can be adjusted if the distributions of \( \boldsymbol{\xi} \) changes.
This increases the accuracy of the PCE over long-term simulations. 

The TD-gPC approach does have some limitations in the application to dynamic systems. For longer simulations, the
accuracy of the method decreases \parencite{gerritsmaTimedependentGeneralizedPolynomial2010}, thus needing recalculation
and adding extra computing cost. \textcite{ebadollahiTimedependentUncertaintyQuantification2025} also mentions a
deterioration of the performance of TD-gPC for longer simulations. The computational cost rapidly increases as the
required polynomial order increases as time evolves. This makes it inefficient for strongly nonlinear dynamic systems.
The technique does not scale well for higher dimension problems
\parencites{heuvelineHybridGeneralizedPolynomial2014}{maiSurrogateModellingStochastic2016}. 

\subsection{Limitations}

Although spectral methods are powerful for problems with smooth responses and low-dimensional uncertainty, they become
impractical for dynamic positioning systems. DP models involve high-dimensional, nonlinear dynamics and stochastic
environmental inputs, which lead to rapid growth in computational cost and poor convergence for spectral approaches.

Additionally, Polynomial Chaos Expansions (PCE) and other spectral decompositions are typically constructed for a
specific set of uncertain parameters and their probability distributions. If these distributions change or new 
parameters are introduced, the expansion must be recomputed or retrained. This means that a tool intended to evaluate
behavior under varying conditions would require repeated recalculation of the PCE, making it inefficient compared to
more flexible approaches such as neural network-based surrogate models. 

\todo{Reformat naming of these sections}
\section{Neural network approaches}

This section outlines existing methods used to capture epistemic and aleatoric uncertainty with neural networks. In
general, approaches capture either epistemic or aleatoric uncertainty seperately \parencite{galUncertaintyDeepLearning2016}. Epistemic uncertainty is modelled by
placing prior distributions over the weights of the model and capturing how these vary given input data. Aleatoric
uncertainty is quantified by placing a distribution over the output of the model \parencite{kendallWhatUncertaintiesWe2017}

\TODO{Decide what to discuss here. I want to talk about different methods used to capture uncertainty commonly used\\
- Bayesian Neural Networks \\
- Monte Carlo Dropout\\
}
\subsection{Bayesian Neural Networks}
\subsection{Monte Carlo Dropout}
\subsection{Bayes by backpropagation}
\subsection{Deep ensembles}
\subsection{Hypermodels}
\subsection{Variational neural networks}
\subsection{Neural SDEs}